A Data-Driven Framework for Neural Field Modelling

### 2010-11-18 ltx/first_order_framework/LaTeX_Manuscript.tex
Response of the help request of section 3.2.3

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modelling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Figure~\ref{fig:Figure1}. To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
Now we shall express the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation}
where $\textrm{D}=d/dt + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretise time using a first-order Euler method (see the supplementary (\dean{Appendix}) material for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. The disturbance is added to account for model uncertainty and unmodelled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

When considering the measurement of a point source in a conductive homogeneous field measured by a point sensor, the observation kernel will theoretically follow a Laplacian like shape, with an exponential decay in the amplitude of a measurement with an increasing distance between the source and sensor~\cite{Jackson1999}. However, when considering a source or sensor with a spatial extent, the peak of the observation kernel becomes smoothed and is more similar to a Gaussian shape~\cite{Jackson1999}. An experiment supporting this was performed and the results are shown in \dean{Appendix?}. Therefore, in this study the observation kernel, $m(\mathbf{r}-\mathbf{r}')$, that governs the sensor pick-up geometry is defined by 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)},
\end{equation}
where $\sigma_m$ sets the kernel width.

@@

The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) over space. The MRMSE of the field estimates over time for 150 realisations is shown in Figure~\ref{fig:Figure9} with the $95~\%$ confidence interval. At each time point, the error is consistently around $0.5$~mV, which is approximately $10.73~\%$ of the field voltage range ([-2.20, 2.46]~mV). The error in the reconstructed field is mostly due to the loss of high spatial frequency information from the basis function decomposition as illustrated in Figure~\ref{fig:Figure10}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field, and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:Figure10}(B) that the estimated field contains the basic structure of the true field, shown in Figure~\ref{fig:Figure10}(A), but is missing the higher-frequency detail, which can be see in the error shown in Figure~\ref{fig:Figure10}(C).

@@

For each of the Mexican hat connectivities, the support of the kernel was assumed to be known. However, in Figure~\ref{fig:Figure8}(D-F) the kernel support in the estimator differed from the actual data. In these cases the data sets were generated using a single positive (excitatory) Gaussian kernel, as seen by the solid line in each of the subplots of the figure. The estimation algorithm was constructed allowing for the Mexican hat connectivity of the previous examples. Each plot shows that the actual kernel lies within the confidence intervals of the reconstructed estimates for the majority of the kernel space, until the amplitude is relatively low. Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Although the net reconstructed kernel estimates are in good accord with the actual kernels, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that unless the support of the kernel is precisely known, the interpretation of the results should be based on the net shape of the estimated connectivity kernel and not on the estimated parameter values individually.

@@

% The size of the squares in the figure show the magnitude of the log of the error. The squares are black if the log of the error is negative (the RMSE is $<1$) and the squares are white if the log of the error is positive (the RMSE is $>1$). The largest black squares indicate the best performance and the largest white squares indicate the worst performance for the parameter combinations (\dean{need some thing more quantitative}). 
From Figure~\ref{fig:Figure12} it can be seen that the estimation accuracy depends on the parameters of the sigmoid function. This is expected since that by altering the parameters we are changing the nonlinearity and the saturation region of the sigmoid.  
\dean{PLEASE HELP TO FILL THIS SECTION IN. ANYTHING WILL DO.}
..\dean{This next bit is CRAP (please help):} When the threshold is negative or the slope is very gentle, all of the activity is being passed through the sigmoid function. However, as the slope is increased or the threshold is increased, more of the field is within the saturating region of the activation function. This is illustrated in Figure~\ref{fig:sigmoidfunctions}, where the sigmoids from the bottom left corner through to near top right corner of Figure~\ref{fig:sigmoidfunctions} are plotted. This appears to have a detrimental effect on the estimation accuracy.

>>>>

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modelling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Figure~\ref{fig:Figure1}. In an attempt model longer range connectivity, To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
To arrive at the final form of the model, we shall express the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation}
where $\textrm{D}=d/dt + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretise time using a first-order Euler method (see the supplementary (\dean{Appendix}) material for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. The disturbance is added to account for model uncertainty and unmodelled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

When considering the measurement of a point source in a conductive homogeneous field measured by a point sensor, the observation kernel will theoretically follow a Laplacian like shape, with an exponential decay in the amplitude of a measurement with an increasing distance between the source and sensor~\cite{Jackson1999}. However, when considering a source or sensor with a spatial extent, the peak of the observation kernel becomes smoothed and is more similar to a Gaussian shape~\cite{Jackson1999}. An experiment supporting this was performed and the results are shown in \dean{Appendix?}. Therefore, in this study the observation kernel, $m(\mathbf{r}-\mathbf{r}')$, that governs the sensor pick-up geometry and is defined by 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)},
\end{equation}
where $\sigma_m$ sets the kernel width.

@@

The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) over space. The MRMSE of the field estimates over time for 150 realisations is shown in Figure~\ref{fig:Figure9} with the $95~\%$ confidence interval. At each time point, the error is consistently around $0.5$~mV, which is approximately $10.73~\%$ of the field voltage range ([-2.20, 2.46]~mV). The error in the reconstructed field is again mostly due to the loss of high spatial frequency information from the basis function decomposition as illustrated in Figure~\ref{fig:Figure10}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field, and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:Figure10}(B) that the estimated field contains the basic structure of the true field, shown in Figure~\ref{fig:Figure10}(A), but is missing the higher-frequency detail, which can be see in the error shown in Figure~\ref{fig:Figure10}(C).

@@

For each of the Mexican hat connectivities, the support of the kernel was assumed to be known. However, in Figure~\ref{fig:Figure8}(D-F) the kernel support in the estimator differed from the actual data. In these cases the data sets were generated using a single positive (excitatory) Gaussian kernel, as seen by the solid line in each of the subplots of the figure. The estimation algorithm was constructed allowing for the Mexican hat connectivity of the previous examples. Each plot shows that the actual kernel lies within the confidence intervals of the reconstructed estimates for the majority of the kernel space, until the amplitude is relatively low. Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Although the net reconstructed kernel estimates are in good accord with the actual kernels, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that unless the widths of the Gaussians forming the kernel are precisely known, the interpretation of the results should be based on the net shape of the estimated connectivity kernel and not on the estimated parameter values individually.

@@

% The size of the squares in the figure show the magnitude of the log of the error. The squares are black if the log of the error is negative (the RMSE is $<1$) and the squares are white if the log of the error is positive (the RMSE is $>1$). The largest black squares indicate the best performance and the largest white squares indicate the worst performance for the parameter combinations (\dean{need some thing more quantitative}). 
From Figure~\ref{fig:Figure12} it can be seen that the estimation accuracy depends on the parameters of the sigmoid function. This is expected since by altering the parameters we are changing the nonlinearity and the saturation region of the sigmoid.

\ken{what I would write}

For example, when the sigmoid slope is very gentle, most of the neural activity falls in the central, nearly linear region of the sigmoid, thus having minor effects on the estimation procedure. On the other hand, sharp slopes result in most of the activity falling in the saturation region of the sigmoid with adverse effects on the estimation. Similar arguments can be made for variations to the sigmoid threshold.

\ken{what do you think?}

\dean{PLEASE HELP TO FILL THIS SECTION IN. ANYTHING WILL DO.}
..\dean{This next bit is CRAP (please help):} When the threshold is negative or the slope is very gentle, all of the activity is being passed through the sigmoid function. However, as the slope is increased or the threshold is increased, more of the field is within the saturating region of the activation function. This is illustrated in Figure~\ref{fig:sigmoidfunctions}, where the sigmoids from the bottom left corner through to near top right corner of Figure~\ref{fig:sigmoidfunctions} are plotted. This appears to have a detrimental effect on the estimation accuracy.
